---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data

```{r}

data <- read.csv("activity.csv",header = TRUE, colClasses = c("numeric","character","integer"))

```



## What is mean total number of steps taken per day?

```{r}

steps_by_day <- aggregate(steps ~ date, data, sum)
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")

rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)



```


The mean is `r as.integer(rmean)`. The median is `r as.integer(rmedian)`.


## What is the average daily activity pattern?

1. Calculate average steps for each interval for all days.
2. Plot the Average Number Steps per Day by Interval.
3. Find interval with most average steps.


```{r}

steps_by_interval <- aggregate(steps ~ interval, data, mean)
plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")

```

```{r}
max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
```

The 5-minute interval, on average across all the days in the data set, containing the maximum number of steps is `r max_interval`.

## Imputing missing values

```{r}

incomplete <- sum(!complete.cases(data))

```

Total number of rows with missing values are `r incomplete`.

```{r}

imputed_data <- transform(data, steps = ifelse(is.na(data$steps), steps_by_interval$steps[match(data$interval, steps_by_interval$interval)], data$steps))
imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0

steps_by_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")

hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="red", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)

rmean.i <- mean(steps_by_day_i$steps)
rmedian.i <- median(steps_by_day_i$steps)

mean_diff <- rmean.i - rmean
med_diff <- rmedian.i - rmedian

total_diff <- sum(steps_by_day_i$steps) - sum(steps_by_day$steps)

```

1. The imputed data mean is `r as.integer(rmean.i)`.
2. The imputed data median is `r as.integer(rmedian.i)`.
3. The difference between the non-imputed mean and imputed mean is `r mean_diff`.
3. The difference between the non-imputed mean and imputed mean is `r med_diff`.
4. The difference between total number of steps between imputed and non-imputed data is `r total_diff`. Thus, there were `r total_diff` more steps in the imputed data.


## Are there differences in activity patterns between weekdays and weekends?

```{r}

weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
head(imputed_data)
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))
head(imputed_data)

steps_by_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)

library(lattice)
xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")

```


